---
title: "Week 2: Workflowr Project Setup"
author: "PSYC 859 Data Management and Visualization"
date: "1/15/2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
editor_options:
  chunk_output_type: console
---

*This assignment is due by 1/28/2026 at 8am.*

## Overview

This assignment builds directly on the workflowr vignette:
<https://workflowr.github.io/workflowr/articles/wflow-01-getting-started.html>.
Your goal is to create a new workflowr project with a clean structure,
publish at least one analysis file, and make it available as a GitHub Pages site.

## Learning goals

- Initialize a workflowr project with a reproducible structure
- Knit analysis files with workflowr and track the build status
- Publish a workflowr site to GitHub Pages

## Setup requirements

Before starting, make sure you have:

- R and RStudio installed
- Git installed and configured
- A GitHub account
- The `workflowr` package installed

```{r}
# Install once if needed
## install.packages("workflowr")
```

## Tasks

### 1. Start a new workflowr project

Create a new project named `workflowr-<your-onyen>` in a location you can find later.

```{r}
library(workflowr)
wflow_start("workflowr-<your-onyen>")
```

Open the project in RStudio and confirm the default folders exist:
`analysis/`, `data/`, `code/`, `output/`, and `docs/`.

```{r}
# Briefly note what you see in the project root.
```

### 2. Create a new analysis file

Create a new analysis file at `analysis/01-getting-started.Rmd`.
Include the following:

- A short paragraph describing the project topic you plan to use this semester
- One simple code chunk that prints a small summary (e.g., `summary(mtcars)`)
- A chunk at the end that runs `sessionInfo()`

```{r}
# Paste the YAML header from your analysis file here.
```

```{r}
# Paste the code chunk that summarizes a dataset here.
```

```{r}
# Paste the sessionInfo() output here.
```

### 3. Build and check the site locally

Use workflowr to build the site and check its status.

```{r}
wflow_build("analysis/01-getting-started.Rmd")
wflow_status()
```

```{r}
# Paste the output of wflow_status() here.
```

### 4. Publish and commit with workflowr

Publish the new analysis page using workflowr so it is tracked by Git.

```{r}
wflow_publish("analysis/01-getting-started.Rmd", message = "Add initial analysis page")
```

```{r}
# Paste the output of wflow_publish() here.
```

### 5. Push to GitHub and enable GitHub Pages

1. Create a new GitHub repository named `workflowr-<your-onyen>`.
2. Add it as a remote and push your commits.
3. In GitHub, enable Pages for the `docs/` folder on the `main` branch.

```{r}
# Paste the GitHub repo URL here.
```

```{r}
# Paste the GitHub Pages URL here.
```

### 6. Verify the published site

Open your GitHub Pages site and confirm:

- The home page loads
- Your analysis page appears in the navigation
- The analysis page renders your text and code output

```{r}
# Briefly describe anything that did not work as expected (if applicable).
```

## Submission checklist

- GitHub repository link
- GitHub Pages link
- `wflow_status()` output
- `wflow_publish()` output
- Short reflection (2-3 sentences) on what was straightforward vs. confusing

```{r}
# Reflection
```
